\begin{summary}
    This project, \textit{Space Oddities}, explores the historical growth and current density of artificial objects in Earth's orbit. By combining metadata from the Union of Concerned Scientists with orbital elements from Space-Track, we aim to visualize the risk of orbital congestion and the emergence of "mega-constellations" like Starlink. Our visualization will guide users from a historical overview of space launches to a visual mapping of collision risks in Low Earth Orbit (LEO).
\end{summary}

\section{Dataset}

\subsection{Data Sources}
To ensure a comprehensive analysis, we will use three primary datasets:
\begin{enumerate}
    \item \textbf{UCS Satellite Database} \cite{ucs-database}: A comprehensive registry detailing active satellites. It provides crucial metadata including ownership, country of origin, purpose, mass, and orbit class.
    \item \textbf{Space-Track.org} \cite{space-track}: Maintained by the US Space Command, this platform provides raw spatial data via Two-Line Elements (TLEs) for the entire catalog of tracked orbital objects, including active payloads and debris.
    \item \textbf{CelesTrak} \cite{celestrak}: Maintained by Dr. T. S. Kelso, this dataset serves as an external source to compare and augment results from Space-Track.
\end{enumerate}


\subsection{Quality \& Preprocessing}
The UCS database is highly structured but was last updated in May 2023, requiring careful synchronization with current orbital data. Space-Track data is updated daily but requires complex parsing of TLE text formats.

\begin{table}[H]
    \centering
    \caption{Dataset Comparison and Characteristics}
    \label{tab:datasets}
    \begin{tabular}{@{}llll@{}} \toprule
        Feature & UCS Database & Space-Track & CelesTrak \\ \midrule
        \textbf{Primary Content} & Active Satellites & TLEs (Active + Debris) & TLEs + SATCAT \\
        \textbf{Format} & Excel / CSV & Text / API & Text / JSON \\
        \textbf{Update Frequency} & Quarterly & Daily & Real-time \\
        \textbf{Preprocessing} & Minimal (Cleaning) & High (TLE Parsing) & Moderate (Validation) \\ \bottomrule
    \end{tabular}
\end{table}

We will use Python libraries such as \texttt{Skyfield} or \texttt{SGP4} to convert raw TLEs into metrics such as altitude, inclination, and semi-major axis. This will also allow us to compare satellite's positions with each other as they orbit around the earth. The primary challenge will be joining metadata from the UCS with spatial data using unique NORAD Catalog IDs, and complementing this information with sources like Celestrak.

\section{Problematic}

\begin{fquote}[Donald J. Kessler][][\cite{kessler-syndrome}]
    The orbit of the Earth is a limited resource. As we place more and more objects into it, the probability of collisions will increase until the collision rate exceeds the rate of removal by atmospheric drag.
\end{fquote}

\subsection{Topic \& Main Axis}
Our visualization explores the exponential growth of orbital objects, focusing on congestion and the threat of the "Kessler Syndrome" \cite{kessler-syndrome}. We aim to answer:
\begin{itemize}
    \item \textbf{Where} is the highest density of objects currently located?
    \item \textbf{How} have mega-constellations altered the landscape of LEO?
    \item \textbf{What} are the current risks of collision ?
\end{itemize}

\begin{invsummary}
    \textbf{The Kessler Syndrome:} A theoretical scenario where the density of objects in LEO is high enough that collisions between objects cause a cascade of debris, potentially making specific orbits unusable for generations.
\end{invsummary}

\subsection{Motivation \& Target Audience}
With the commercial spaceflight boom, space sustainability has become a critical global issue. Our target audience includes the general public and policymakers who require an intuitive understanding of orbital density that raw data alone cannot provide.

\section{Exploratory Data Analysis}
\subsection{Initial Statistics}
Initial exploration of the UCS dataset confirms a massive spike in LEO satellites over the last five years, largely driven by commercial telecommunications. We will generate the following visualizations for our initial analysis:
\begin{itemize}
    \item \textbf{Time-series Analysis:} Total objects launched per year versus objects currently in orbit.
    \item \textbf{Mass Distribution:} Categorizing mass by country and purpose (e.g., Commercial vs. Military).
    \item \textbf{Orbital Mapping:} Plotting density as a function of altitude (per km) to identify peak congestion zones.
\end{itemize}

\subsection{Insights}
Preliminary data suggests that the region between 500km and 600km altitude contains the highest density of active payloads. Furthermore, the Space-Track catalog reveals that fragmentation events (anti-satellite tests, accidental collisions) account for the vast majority of trackable debris, significantly outweighing active payloads.

\section{Related work}
\subsection{Existing Visualization Projects}
\begin{itemize}
    \item \textbf{Stuff in Space} \cite{stuff-in-space}: A 3D WebGL visualization showing real-time positions. While technically impressive, it lacks a guided narrative.
    \item \textbf{LeoLabs} \cite{leolabs}: A commercial platform for conjunction monitoring. However, it is primarily a proprietary tool for satellite operators.
\end{itemize}

\subsection{Originality \& Inspiration}
Our approach differs by being \textbf{narrative-driven}. We will not just plot 3D points; we will integrate 2D temporal charts with a filtered 3D view to visualize "orbital highways." By incorporating risk-based data (CDMs), we uniquely visualize the \textit{potential} for disaster rather than just the \textit{presence} of objects. We draw inspiration from the aesthetic of \textit{Stuff in Space} but intend to provide the data depth of commercial tools like LeoLabs in a publicly accessible, educational format.
